---
title: "Analysis of CARNIVAL results"
author: "Alberto Valdeolivas: alberto.valdeolivas@bioquant.uni-heidelberg.de; Date:"
date: "16/07/2020"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### License Info

This program is free software: you can redistribute it and/or modify it under 
the terms of the GNU General Public License as published by the Free Software 
Foundation, either version 3 of the License, or (at your option) any later 
version.

This program is distributed in the hope that it will be useful, but WITHOUT ANY 
WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR 
A PARTICULAR PURPOSE. See the GNU General Public License for more details.

Please check http://www.gnu.org/licenses/.

## Introduction

The present script takes the CARNIVAL results inferred from transcriptomics data 
from intestinal organoids treated with Sars-CoV-2 for 60 hours and makes some 
enrichment analysis and clustering. 

The original transcriptomic dataset is coming from the following publication:
[Lamers et al. 2020](https://science.sciencemag.org/content/369/6499/50).

The differential expression analysis was conducted by Martina Poletti (<Martina.Poletti@earlham.ac.uk>) 

## Getting Ready 

### Libraries and functions

We first load the required libraries and some important functions to analyse 
CARNiVAL output. 

```{r, warning=FALSE, message=FALSE}
library(readr)
library(fgsea)
library(dplyr)
library(ggplot2)
library(piano)
library(tidyr)
library(ggsci)
library(igraph)
library(gprofiler2)
library(plotly)

## Function to extract the nodes that appear in CARNIVAL network and the 
## background genes (all genes present in the prior knowledge network).
## It returns a list with two objects: the success and the background genes.
extractCARNIVALnodes <- function(CarnivalResults){

    CarnivalNetwork <- 
        as.data.frame(CarnivalResults$weightedSIF, stringsAsFactors = FALSE)
    
    colnames(CarnivalNetwork) <- c("source", "sign", "target", "Weight")

    ## We define the set of nodes interesting for our condition
    sucesses <- unique(c(gsub("_.*","",CarnivalNetwork$source), 
        gsub("_.*","",CarnivalNetwork$target)))

    CarnivalAttributes <- as.data.frame(CarnivalResults$nodesAttributes, 
        stringsAsFactors = FALSE)

    ## We define the background as all the genes in our prior knowledge network.
    bg <- unique(gsub("_.*","",CarnivalAttributes$Node))     
    
    return(list(sucesses = sucesses, bg= bg))
}

### Function to print a barplot with the enriched pathways.
BarplotEnrichment <- function(PathwaysSelect, Interesting_pathways){ 
    
    p <- ggplot(PathwaysSelect, aes(x = reorder(pathway, pvalue), 
            y = -log10(pvalue))) + 
        geom_bar(aes(fill = mean_stat), stat = "identity") +
        scale_fill_gradient2(low = "darkblue", high = "indianred", 
            mid = "whitesmoke", midpoint = 0) + 
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 90, hjust = 1, 
            colour = ifelse(levels(reorder(PathwaysSelect$pathway, 
                PathwaysSelect$pvalue)) %in% Interesting_pathways, 
                "red", "grey40"),
            face = ifelse(levels(reorder(PathwaysSelect$pathway, 
                PathwaysSelect$pvalue)) %in% Interesting_pathways, 
                "bold", "plain")),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) + 
        xlab("")
    return(p)
}    
```

### Reading CARNIVAL results

We then read the CARNIVAL results generated in the previous script. We define 
two different gene sets in order tor conduct the enrichment. The first set 
contains the nodes that appear in the CARNIVAL output and are therefore 
relevant in the context of our input transcriptomic data. The second set 
contains all the genes in our prior knowledge network which are used as the 
backgroud.

```{r}
carnival_results_inhibition <- 
    readRDS("Carnival_Results/carnival_results_top50tf_pleitropic_minsize15.rds")
nodes_carnival_inhibition <- extractCARNIVALnodes(carnival_results_inhibition)

carnival_results_activation <- 
    readRDS("Carnival_Results/carnival_results_top50tf_pleitropic_minsize15_activation.rds")
nodes_carnival_activation <- extractCARNIVALnodes(carnival_results_activation)

carnival_results_undefined <- 
    readRDS("Carnival_Results/carnival_results_top50tf_pleitropic_minsize15_undefinedEffect.rds")
nodes_carnival_undefined <- extractCARNIVALnodes(carnival_results_undefined)
```

### Reading Pathway data sets from MSigDB

We downloaded from MSigDB https://www.gsea-msigdb.org/ the following 
dataset: c2.cp.v7.0.symbols.gmt. It contains several pathways from different 
resources and the genes that are known to be involved in those pathways. 

```{r}
Pathway_signatures <- loadGSC("InputData/c2.cp.v7.0.symbols.gmt")
```

### Reading and formatting statistic from DEG

We read the results from the differential expression analysis. The statistic of 
the genes will be mapped later on in the different significant pathways. Maybe,
this is not very informative in this context. I am going to also check with the
Average node activity from CARNIVAL. 
    
```{r}
DEA_results <- read_csv("InputData/Clevers_degs_all.csv") %>%
   dplyr::select(Gene, stat)

mean_stat <- unlist(lapply(Pathway_signatures$gsc, function(x, DEA_results) {
    genes_matching <- x[which(x %in% DEA_results$Gene)]
    mean_genes <- dplyr::filter(DEA_results, Gene %in% genes_matching) %>%
        dplyr::pull(stat) %>% mean(na.rm = TRUE)
    return(mean_genes)
}, DEA_results = DEA_results)) %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "pathway") %>%
    dplyr::rename(mean_stat = ".")

### Trying with Carnival node activity
carnival_avg_node_act <- carnival_results_inhibition$nodesAttributes %>% 
    as.data.frame() %>%
    dplyr::mutate(AvgAct = as.numeric(AvgAct))

mean_stat_carni <- 
    unlist(lapply(Pathway_signatures$gsc, function(x, carnival_avg_node_act) {
    genes_matching <- x[which(x %in% carnival_avg_node_act$Node)]
    mean_genes <- dplyr::filter(carnival_avg_node_act, Node %in% genes_matching) %>%
        dplyr::pull(AvgAct) %>% mean(na.rm = TRUE)
    return(mean_genes)
}, carnival_avg_node_act = carnival_avg_node_act)) %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "pathway") %>%
    dplyr::rename(mean_stat = ".")
```

### Performing Enrichment Analysis and plotting the Results

Using the Piano R package, we run a gene set analysis (GSA) based on a list of 
significant genes (CARNIVAL nodes) and a gene set collection (background). 
It uses Fisher’s exact test.

First for CARNIVAL results when we assume that the viral-host proteins 
interactions are inhibitory: 

```{r}
hyper_results_inhibition <- 
    runGSAhyper(genes = nodes_carnival_inhibition$sucesses, 
        # pvalues = rep(0,length(nodes_carnival_inhibition$sucesses)),
        pcutoff = 0.01, 
        universe = nodes_carnival_inhibition$bg, gsc = Pathway_signatures,
        gsSizeLim = c(4,Inf), adjMethod = "fdr")

enriched_pathways_inhibition <- hyper_results_inhibition$resTab %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "pathway") %>% 
    as_tibble() %>% 
    dplyr::filter(`Adjusted p-value` <= 0.01) %>% 
    dplyr::inner_join(mean_stat_carni) %>% 
    dplyr::select(pathway, `p-value`, `Adjusted p-value`, mean_stat) %>%
    dplyr::rename(pvalue = `p-value`, AdjPvalu = `Adjusted p-value`) %>% 
    dplyr::mutate(pathway = as.factor(pathway))
```

```{r,dpi=300, warning=FALSE, message=FALSE}
interesting_pathways_inhibition <- c()

p_inhibition <- BarplotEnrichment(enriched_pathways_inhibition, 
    interesting_pathways_inhibition)
```

```{r, fig.width=12, fig.height=12, echo=FALSE, dip=300}
p_inhibition
```

Then, for CARNIVAL when we assume that the viral-host proteins interactions are 
stimulations: 

```{r}
hyper_results_activation <- 
    runGSAhyper(genes = nodes_carnival_activation$sucesses, 
        # pvalues = rep(0,length(nodes_carnival_activation$sucesses)),
        pcutoff = 0.01, 
        universe = nodes_carnival_activation$bg, gsc = Pathway_signatures,
        gsSizeLim = c(4,Inf), adjMethod = "fdr")

enriched_pathways_activation <- hyper_results_activation$resTab %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "pathway") %>% 
    as_tibble() %>% 
    dplyr::filter(`Adjusted p-value` <= 0.01) %>% 
    dplyr::inner_join(mean_stat_carni) %>% 
    dplyr::select(pathway, `p-value`, `Adjusted p-value`, mean_stat) %>%
    dplyr::rename(pvalue = `p-value`, AdjPvalu = `Adjusted p-value`) %>% 
    dplyr::mutate(pathway = as.factor(pathway))
```

```{r,dpi=300, warning=FALSE, message=FALSE}
interesting_pathways_activation <- c()

p_activation <- BarplotEnrichment(enriched_pathways_activation, 
    interesting_pathways_activation)
```

```{r, fig.width=12, fig.height=12, echo=FALSE, dip=300}
p_activation
```

### Comparing Enrichment between activation and inhibition

Now I compare the enriched pathways that change the most between both 
assumptions: host-viral interactions are inhibitory or stimulatory. 

```{r}
results_activation <-   hyper_results_activation$resTab %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "pathway") %>% 
    as_tibble() %>% 
    dplyr::mutate(LogpValue_activation = -log(`p-value`), 
        p_value_activation = `p-value`) %>% 
    dplyr::select(pathway, p_value_activation, LogpValue_activation) 

results_inhibition <-   hyper_results_inhibition$resTab %>% 
    as.data.frame() %>% 
    tibble::rownames_to_column(var = "pathway") %>% 
    as_tibble() %>% 
    dplyr::mutate(LogpValue_inhibition = -log(`p-value`), 
        p_value_inhibition = `p-value`) %>% 
    dplyr::select(pathway, p_value_inhibition, LogpValue_inhibition) 

results_join <- 
    dplyr::inner_join(results_activation, results_inhibition) %>% 
    dplyr::filter(p_value_activation < 0.05 | p_value_inhibition < 0.05) %>% 
    dplyr::mutate(diff = LogpValue_activation - LogpValue_inhibition) %>% 
    dplyr::top_n(15, wt = abs(diff)) %>% 
    dplyr::arrange(desc(abs(diff)))

results_join_longer <- results_join %>% 
    dplyr::select(pathway, diff, LogpValue_activation, LogpValue_inhibition) %>% 
    dplyr::rename(activation = "LogpValue_activation", 
        inhibition = "LogpValue_inhibition") %>% 
    pivot_longer(-c(pathway,diff), values_to = "LogPvalue", 
        names_to = "Interaction")  
    
point_plot_activation_vs_inhibtion <- 
    ggplot(results_join_longer, aes(reorder(pathway, abs(diff)), LogPvalue)) + 
    geom_point(aes(color = Interaction), size = 3) + 
    coord_flip() + 
    theme_minimal() + 
    theme(legend.position = "bottom",  legend.justification = "center") +
    scale_color_lancet() +
    theme(axis.text.x = element_text(hjust = 1, size =8.5),
        axis.text.y = element_text(size = 7),
        panel.grid.minor = element_blank()) + 
    xlab("Pathways") + ylab("-Log (p-value)") +
    geom_hline(yintercept = -log(0.05), linetype="dashed", 
                color = "#2F4F4F", size=0.5)
```

```{r, dpi=300}
point_plot_activation_vs_inhibtion
```

### Clustering of CARNIVAL solution and enrichment. 

We first convert our CARNIVAL output network into an igraph object. Then, we 
perform different types of clustering and select the one with the largest 
modularity for further analysis. 

```{r, warning=FALSE,message=FALSE}
carnival_results_igraph <- carnival_results_undefined$weightedSIF %>%
    as.data.frame() %>% dplyr::select(-Sign) %>% 
    dplyr::rename(weight = "Weight") %>% 
    igraph::graph_from_data_frame(directed = FALSE)

c_edgeBetweeness <- 
    cluster_edge_betweenness(carnival_results_igraph)

c_FastGreedy <- 
    cluster_fast_greedy(carnival_results_igraph, 
    merges = TRUE, 
    modularity = TRUE, 
    membership = TRUE, 
    weights = E(carnival_results_igraph)$weight)

c_Infomap <- 
    cluster_infomap(carnival_results_igraph, 
    e.weights =  E(carnival_results_igraph)$weight, 
    v.weights = NULL, nb.trials = 10, modularity = TRUE)

c_labelProp <- 
    cluster_label_prop(carnival_results_igraph, 
    weights = E(carnival_results_igraph)$weight)

c_leadingEig <- 
    cluster_leading_eigen(carnival_results_igraph, 
    weights = E(carnival_results_igraph)$weight)

c_Louvain<- 
    cluster_louvain(carnival_results_igraph, 
    weights = E(carnival_results_igraph)$weight)

c_walktrap <- 
    cluster_walktrap(carnival_results_igraph, 
    weights = E(carnival_results_igraph)$weight, 
    steps = 4, merges = TRUE, modularity = TRUE, membership = TRUE)

modularity_df <- data.frame(method = c("EdgeBetweeness", "FastGreedy", 
    "Infomap","LabelPropagation","LeadingEigenvector", "Louvain","Walktrap"), 
    modularity = c(max(c_edgeBetweeness$modularity), 
                   max(c_FastGreedy$modularity), 
                   max(c_Infomap$modularity), 
                   max(c_labelProp$modularity),
                   max(c_leadingEig$modularity),
                   max(c_Louvain$modularity),
                   max(c_walktrap$modularity))) %>% 
    dplyr::arrange(desc(modularity))
modularity_df
# karate <- make_graph("Zachary")
# wc <- cluster_walktrap(karate)
# modularity(wc)
# membership(wc)
# plot(wc, karate)
```

We then selected the communities generated by the Fast Greedy method. We will 
perform enrichment analysis on them. 

```{r, warning=FALSE, message=FALSE}
## Number of genes in the different communities. 
table(c_FastGreedy$membership)

## We perform an Enrichment Analysis for each Community: 
n <- length(unique(c_FastGreedy$membership))
gene_communities <- 
    data.frame(Gene = c_FastGreedy$names, Community = c_FastGreedy$membership)
p <- list()

for (i in seq_len(n)){
    current_genes <- dplyr::filter(gene_communities, Community==i) %>%
        dplyr::pull(Gene)
    current_result <- gost(current_genes, user_threshold = 0.01, 
        correction_method = "fdr", custom_bg = nodes_carnival_undefined$bg,
        sources = c("GO","KEGG","REAC","WP")) 
    if (!is.null(current_result)){
        p[[i]] <- gostplot(current_result, capped = FALSE, interactive = TRUE)
    }
}
```

```{r}
p[[1]]
```


## Session Info Details

```{r, echo=FALSE, eval=TRUE}
sessionInfo()
```
